Equivalence Testing and the Second Generation \emph{P}-Value

### 2019-04-22 reply_to_reviews.R2.tex

\linenumbers

\authornote{

Correspondence concerning this article should be addressed to Daniël Lakens, Den Dolech 1, IPO 1.33, 5600 MB, Eindhoven, The Netherlands. E-mail: \href{mailto:D.Lakens@tue.nl}{\nolinkurl{D.Lakens@tue.nl}}}

\abstract{
In this second revision we are responding to the last issue you raised, about our incorrect statement that 50\% of the correlations fall below the observed correlation. We are extrmely grateful you pointed out this incorrect statement, and have corrected it as explained below. We hope you think our submission is ready to be accepted for publication in Meta-Psychology.

@@

}

\begin{document}
\maketitle

\textbf{Dear editor,}

\textbf{thank you for your comments on our revised manuscript. You rightly noted that our statement that \enquote{There is always a 50\% probability of observing a correlation smaller or larger than the true correlation} was not correct. Our statement was true for the corresponding z-value when transforming the correlation while calculating confidence intervals, but not for correlations, which are not symmetrically distributed. We have deleted the following:}

\begin{quote}
\emph{In other words, if the true effect size is the same as the equivalence bound, it is equally likely to find an effect more extreme than the equivalence bound, as it is to observe an effect that is less extreme than the equivalence bound.}
\end{quote}

\textbf{We have made a number of small changes. First, we clearly explain that, although the z-distribution is symmetrical, the correlation is not. We note how the transformation has the goal to yield accurate error rates (which is important in the TOST procedure) on page 18, lines 250-255.}

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

If the observed correlation falls exactly on the equivalence bound the \emph{p}-value for the equivalence test is 0.5. In the equivalence test for correlations the \emph{p} value is computed based on a z-transformation which better controls error rates (Goertzen \& Cribbie, 2010). This transformation is computed as follows, where \emph{r} is the observed correlation and \(\rho\) is the theoretical correlation under the null:

\[
z = \frac{\frac{\log(\frac{1 + r}{ 1 - r})}{2} - \frac{\log(\frac{1 + \rho}{ 1 - \rho})}{2}}{\sqrt{\frac{1}{n-3}}}
\]

Because the z-distribution is symmetric, the probability of observing the observed or more extreme z-score, assuming the equivalence bound is the true effect size, is 50\%. However, because the \emph{r} distribution is not symmetric, this does not mean that there is always a 50\% probability of observing a correlation smaller or larger than the true correlation.

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

\textbf{Second, we have added an explicit statement that in the extreme case, the probability of observing a correlation smaller than the equivalence bound is not 50\%, but in our example, 36\% on page 19, lines 271-275.}

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to \emph{r} = -0.99 and 0.99, with a true correlation of 0.99) 97.60\% of the confidence interval overlaps with the equivalence range, even though in the long run only 36\% of the correlations observed in the future will fall in this range.

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

\textbf{We again want to sincerely thank you for catching this error before our paper was publlished.}

@@

\newpage

\hypertarget{references}{%
\section{References}\label{references}}

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-goertzen_detecting_2010}{}%
Goertzen, J. R., \& Cribbie, R. A. (2010). Detecting a lack of association: An equivalence testing approach. \emph{British Journal of Mathematical and Statistical Psychology}, \emph{63}(3), 527--537. doi:\href{https://doi.org/10.1348/000711009X475853}{10.1348/000711009X475853}

### 2019-04-22 manuscript.R2.Rmd
reply to reviews 2

#The relationship between *p*-values from TOST and SGPV when confidence intervals are symmetrical

@@

##SGPV as a uniform measure of overlap

@@

##When different *p*-values for equivalence tests yield the same SGPV

@@

```{r, TOSTSGPV10, fig.cap="The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure \\ref{fig:TOSTSGPV9}."}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```

@@

##The relation between equivalence tests and SGPV for asymmetrical confidence intervals around correlations

So far we have only looked at the relation between equivalence tests and the SGPV when confidence intervals are symmetric (e.g., for confidence intervals around mean differences). For correlations, which are bound between -1 and 1, confidence intervals are only symmetric for a correlation of exactly 0. The confidence interval for a correlation becomes increasingly asymmetric as the observed correlation nears -1 or 1. For example, with ten observations, an observed correlation of 0 has a symmetric 95% confidence interval ranging from -0.630 to 0.630, while and observed correlation of 0.7 has an asymmetric 95% confidence interval ranging from 0.126 to 0.993. Note that calculating confidence intervals for a correlation involves a Fisher's z-transformation, which transforms values such that they are approximately normally z-distributed, which allows one to compute symmetric confidence intervals. These confidence intervals are then retransformed into a correlation, where the confidence intervals are asymmetric if the correlation is not exactly zero.

@@

```{r, include=FALSE}
low_eqbound_r = -0.45
high_eqbound_r = 0.45
res<-TOSTr(n = 30,
           r = 0.45,
           low_eqbound_r = -0.45,
           high_eqbound_r = 0.45,
           alpha = .05,
           plot = FALSE,
           verbose = FALSE)
sgpv <- p_delta(res$LL_CI_TTEST,res$UL_CI_TTEST, low_eqbound_r, high_eqbound_r)
```

The reason that the equivalence test and SGPV no longer overlap is also because of asymmetric confidence intervals. If the observed correlation falls exactly on the equivalence bound the *p*-value for the equivalence test indicates that the probability of observing the observed or more extreme data, assuming the equivalence bound is the true effect size, is 50%. In other words, if the true effect size is the same as the equivalence bound, it is equally likely to find an effect more extreme than the equivalence bound, as it is to observe an effect that is less extreme than the equivalence bound. However, as can be seen in Figure \ref{fig:TOSTSGPV12}, the two second generation *p*-values associated with the observed correlations at *r* = -0.45 and *r* = 0.45 are larger than 50%. Because the confidence intervals are asymmetric around the observed effect size of 0.45 (ranging from `r round(res$LL_CI_TTEST,2)` to `r round(res$UL_CI_TTEST,2)`) according to @blume_second-generation_2018 `r round(sgpv*100,3)`% of the data-supported hypotheses are null hypotheses, and therefore `r round(sgpv,3)*100`% of the data-supported hypotheses are compatible with the null premise.

@@

This example illustrates the difference between a proportion and a probability. As long as data are sampled from a bivariate normal distribution, there is always a 50% probability of observing a correlation smaller or larger than the true correlation, but the SGPV for this situation depends on how far away the observed correlation is from 0. The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to r = -0.99 and 0.99, with an observed correlation of 0.99) `r round(sgpv,3)*100`% of the confidence interval overlaps with the equivalence range, even though in the long run only 50% of the correlations observed in the future will fall in this range. It should be noted that in larger sample sizes the SGPV is closer to 0.5 whenever the observed correlation falls on the equivalence bound, but this extreme example nevertheless clearly illustrates the difference between question the SGPV answers, and the question a *p*-value answers. The conclusion of this section on asymmetric confidence intervals is that a SGPV of 1 or 0 can still be interpreted as a *p* < 0.025 or *p* > 0.975 in an equivalence test, since the SGPV and *p*-value for the TOST procedure are always directly related at the values *p* = 0.025 and *p* = 0.975. Although @blume_second-generation_2018 state that "the degree of overlap conveys how compatible the data are with the null premise" this definition of what the SGPV provides does not hold for asymmetric confidence intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as 'compatibility with the null hypothesis' under the assumption of a bivariate normal distribution, and the generalizability of this statement needs to be examined beyond normal bivariate distributions. Indeed, Blume and colleagues write in the supplemental material that "The magnitude of an inconclusive second-generation *p*-value can vary slightly when the effect size scale is transformed. However definitive findings, i.e. a *p*-value of 0 or 1 are *not* affected by the scale changes."

@@

##What are the Relative Strengths and Weaknesses of Equivalence Testing and the SGPV?

@@

#Conclusion

>>>>

# The relationship between *p*-values from TOST and SGPV when confidence intervals are symmetrical

@@

## SGPV as a uniform measure of overlap

@@

## When different *p*-values for equivalence tests yield the same SGPV

@@

```{r, TOSTSGPV10, fig.cap="The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure \\ref{fig:TOSTSGPV9}."}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")

@@

## The relation between equivalence tests and SGPV for asymmetrical confidence intervals around correlations

So far we have only looked at the relation between equivalence tests and the SGPV when confidence intervals are symmetric (e.g., for confidence intervals around mean differences). For correlations, which are bound between -1 and 1, confidence intervals are only symmetric for a correlation of exactly 0. The confidence interval for a correlation becomes increasingly asymmetric as the observed correlation nears -1 or 1. For example, with ten observations, an observed correlation of 0 has a symmetric 95% confidence interval ranging from `r round(TOSTr(r = 0, n = 10, low_eqbound_r = -0.45, high_eqbound_r = 0.45, alpha = 0.05, plot = FALSE, verbose = FALSE)$LL_CI_TTEST,3)` to `r round(TOSTr(r = 0, n = 10, low_eqbound_r = -0.45, high_eqbound_r = 0.45, alpha = 0.05, plot = FALSE, verbose = FALSE)$UL_CI_TTEST,3)`, while and observed correlation of 0.7 has an asymmetric 95% confidence interval ranging from `r round(TOSTr(r = 0.7, n = 10, low_eqbound_r = -0.45, high_eqbound_r = 0.45, alpha = 0.05, plot = FALSE, verbose = FALSE)$LL_CI_TTEST,3)` to `r round(TOSTr(r = 0.7, n = 10, low_eqbound_r = -0.45, high_eqbound_r = 0.45, alpha = 0.05, plot = FALSE, verbose = FALSE)$UL_CI_TTEST,3)`. Note that calculating confidence intervals for a correlation involves a Fisher's z-transformation, which transforms values such that they are approximately normally z-distributed, which allows one to compute symmetric confidence intervals. These confidence intervals are then retransformed into a correlation, where the confidence intervals are asymmetric if the correlation is not exactly zero.

@@

```{r, include=FALSE}
low_eqbound_r = -0.45
high_eqbound_r = 0.45
res_1<-TOSTr(n = 30,
           r = 0.45,
           low_eqbound_r = -0.45,
           high_eqbound_r = 0.45,
           alpha = .05,
           plot = FALSE,
           verbose = FALSE)
sgpv <- p_delta(res_1$LL_CI_TTEST,res_1$UL_CI_TTEST, low_eqbound_r, high_eqbound_r)

low_eqbound_r = -0.99
high_eqbound_r = 0.99
res_2<-TOSTr(n = 4,
           r = 0.99,
           low_eqbound_r = low_eqbound_r,
           high_eqbound_r = high_eqbound_r,
           alpha = .05,
           plot = FALSE,
           verbose = FALSE)
sgpv2 <- p_delta(res_2$LL_CI_TTEST,res_2$UL_CI_TTEST, low_eqbound_r, high_eqbound_r)

The reason that the equivalence test and SGPV no longer overlap is due to asymmetric confidence intervals. If the observed correlation falls exactly on the equivalence bound the *p*-value for the equivalence test is 0.5. In the equivalence test for correlations the *p* value is computed based on a z-transformation which better controls error rates [@goertzen_detecting_2010]. This transformation is computed as follows, where  *r* is the observed correlation and $\rho$ is the theoretical correlation under the null:

$$
z = \frac{\frac{\log(\frac{1 + r}{ 1 - r})}{2} - \frac{\log(\frac{1 + \rho}{ 1 - \rho})}{2}}{\sqrt{\frac{1}{n-3}}}
$$

Because the z-distribution is symmetric, the probability of observing the observed or more extreme z-score, assuming the equivalence bound is the true effect size, is 50%. However, because the *r* distribution is not symmetric, this does not mean that there is always a 50% probability of observing a correlation smaller or larger than the true correlation. As can be seen in Figure \ref{fig:TOSTSGPV12}, the two second generation *p*-values associated with the observed correlations at *r* = -0.45 and *r* = 0.45 are larger than 50%. Because the confidence intervals are asymmetric around the observed effect size of 0.45 (ranging from `r round(res_1$LL_CI_TTEST,2)` to `r round(res_1$UL_CI_TTEST,2)`) according to @blume_second-generation_2018 `r round(sgpv*100,3)`% of the data-supported hypotheses are null hypotheses, and therefore `r round(sgpv*100,3)`% of the data-supported hypotheses are compatible with the null premise.

@@

This example illustrates the difference between a proportion and a probability. As long as data are sampled from a bivariate normal distribution, there is always a 50% probability of observing a correlation smaller or larger than the true correlation, but the SGPV for this situation depends on how far away the observed correlation is from 0. The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to *r* = -0.99 and 0.99, with a true correlation of 0.99) `r round(sgpv2,3)*100`% of the confidence interval overlaps with the equivalence range, even though in the long run only 36% of the correlations observed in the future will fall in this range.

It should be noted that in larger sample sizes the SGPV is closer to 0.5 whenever the observed correlation falls on the equivalence bound, but this extreme example nevertheless clearly illustrates the difference between question the SGPV answers, and the question a *p*-value answers. The conclusion of this section on asymmetric confidence intervals is that a SGPV of 1 or 0 can still be interpreted as a *p* < 0.025 or *p* > 0.975 in an equivalence test, since the SGPV and *p*-value for the TOST procedure are always directly related at the values *p* = 0.025 and *p* = 0.975. Although @blume_second-generation_2018 state that "the degree of overlap conveys how compatible the data are with the null premise" this definition of what the SGPV provides does not hold for asymmetric confidence intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as 'compatibility with the null hypothesis' under the assumption of a bivariate normal distribution, and the generalizability of this statement needs to be examined beyond normal bivariate distributions. Indeed, Blume and colleagues write in the supplemental material that "The magnitude of an inconclusive second-generation *p*-value can vary slightly when the effect size scale is transformed. However definitive findings, i.e. a *p*-value of 0 or 1 are *not* affected by the scale changes."

@@

## What are the Relative Strengths and Weaknesses of Equivalence Testing and the SGPV?

@@

# Conclusion
